Real-time Attention Assessment for Social Robots

### 2015-09-17 paper.tex
Shuffle around pieces of text, new figure to illustrate matching between ground truth and computed VFoA

@@

The current state-of-the-art techniques found in HRI that operate both on-line
and during field experiments rely on head pose
estimation, most often based on depth sensors (RGB-D). Fanelli~\etal provides a
overview of these approaches in~\cite{fanelli2012real}. While fairly robust, these
techniques remain CPU intensive for a limited accuracy.

In our work, we rely on recent advances on template-based face
alignment~\cite{kazemi2014one} that allows fast (in the order of a few
milliseconds) face features extraction on 2D images, that we found to be fairly
robust to real-world lighting conditions and occlusions naturally occurring in
child-robot interactions.

In this way, in situ Visual Focus of Attention
(VFoA) estimation is advantageous to human-robot interaction and a major
contribution towards a better understanding of the interaction quality.

\cite{kennedy2015head} -> it does not work!

@@

In \cite{peters2010investigating} approach, a model of user's interest and
engagement with a virtual agent displaying shared attention behavior, by using
eye gaze and head direction information is proposed. Similarly, in
\cite{nakano2010estimating} an approach to estimate the user's conversational
engagement with a conversational agent based on analysis of gaze patterns was
implemented. Other significant work \cite{Castellano:2009} focus on feature
evaluation related to the user's non-verbal behavior, the task and the
companion's affective reactions are identified to predict the children's level
of engagement using a Bayesian model.

@@

\fixme{Litterature review on visual attention...?}

According to \cite{stiefelhagen2002tracking} the head orientation's contribution
in overall gaze direction is 68.9\% showing agreement between both measurements.
In addition, it shows that the head orientation alone can get an average
accuracy of 88.7\% detecting the focus of attention in a meeting application
scenario using a head-mounted display with eye and head tracking.

@@

, to detect and extract the skeleton
of the subject's face and then, use the obtained features to calculate the
head pose orientation estimation (see figure \ref{system}). In this way, we can
estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
representation as it is executed as a ROS (Robot Operating System) node that can
be seen in figure \ref{rviz}. Compared to intrusive hardware like head mounted
displays, eye-trackers or wearables for instance, this approach has a clear
advantage of simplicity and integration with robotic systems, but also becomes
accurate enough for general purposes in face-to-face interactions. In addition,
it allows us to model complex environments, handling different number and
positions of participants in the scene.

\paragraph{Setting the \textit{tf} Frame}\fixme{tf is ROS specific: we should
not write the article in a ROS specific way}
The second step is to use the matrix [R|t] to set the correct translation and
orientation of the \textit{tf} frame which represents the face in the 3D space.
In order to achieve that it is necessary first, to set the origin defined by the
vector \textit{t}. Second, the matrix \textit{R} needs to be reformulated into a
quaternion as shown in equation \ref{eq:quat} due to the smoothly interpolation
between them.

\begin{equation}
\begin{bmatrix}
qw \\
qx \\
qy \\
qz
\end{bmatrix}
=
\begin{bmatrix}
\sqrt{1 + r_{11} + r_{22} + r_{33}} /2 \\
(r_{32} - r_{23})/( 4 \cdot qw) \\
(r_{13} - r_{31})/( 4 \cdot qw) \\
(r_{21} - r_{12})/( 4 \cdot qw)
\end{bmatrix}
\label{eq:quat}
\end{equation}
\\
However, it is necessary to consider the cases where the division is performed
by 0 (in 180\degree rotation about the y-axis for instance) or when \textit{qw}
becomes close to zero.

\paragraph{Computing the Field of Attention}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\columnwidth]{drawSetup}
    \caption{\small Graphical representation of the set-up and the possible children focuses of attention in the scene, in red.}
    \label{drawSetup}
\end{figure}

@@

@@

Our results for a peer-to-peer work scenarios using an approach based on RGB
camera are summarized in table \ref{tab:results} where the \textit{expected} is
the focus of attention that given the state of the system, it is likely to
attract the subject's gaze. In the same manner, the \textit{captured} is the
focus of attention our approach provides and the \textit{real}, the ground truth
of the subject's gaze direction based on manual annotations from two evaluators
with an average reliability of $ \alpha = .89 $.

In order to calculate the inter-rater reliability between the two measurements
(real-captured and real-expected), the overlap percentage between the two curves
has been computed. A more reliable measurement in terms of time shifting between
the real and captured focuses of attention is provided using the cosine
similarity also shown in table \ref{tab:results}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.8\columnwidth]{realExpected}
    \caption{\small In blue, real VFoA of subject 4 accoding to two observer's
    assessment. In cyan, the matching between the real and the obtained VFoA. In
    yellow, the repeated pattern in turn-taking and in orange, the experimenter's
    suggestion to choose a new word.}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Looking at techniques that both operate on-line and have been deployed in field
experiments, one finds that most approaches rely on head pose estimation alone
(no eye gaze tracking) and are most often based on depth sensors (RGB-D) .
Fanelli~\etal provides a overview of these approaches in~\cite{fanelli2012real},
and recent examples include~\cite{baxter2014tracking} or~\cite{anzalone}.

Approaches based on monocular 2D vision have been explored as
well~\cite{peters2010investigating}, with however limited robustness to
occlusions or lightning conditions, and over-reliance on tracking to maintain
real-time performances.  Our work relies on recent advances on template-based
face alignment~\cite{kazemi2014one} that allows fast (in the order of a few
milliseconds) facial features extraction on 2D images, combined with 3D model
fitting, to obtain a fast, robust and stable 6D head pose estimate, that was
successfully deployed in field experiments involving child-robot interactions.

We then derive the field of attention from the head pose: according
to~\cite{stiefelhagen2002tracking} the head orientation's contribution in
overall gaze direction is 68.9\% showing agreement between both measurements.
In addition, it shows that the head orientation alone can get an average
accuracy of 88.7\% detecting the focus of attention in a meeting application
scenario using a head-mounted display with eye and head tracking.

While previous preliminary research in HRI seemed to indicate that deriving
attentional focus from head pose alone would not be accurate
enough~\cite{kennedy2015head}, we found in our case acceptable levels of
agreement between the robot observation and manual post-hoc annotations.

@@

In~\cite{peters2010investigating} approach, a model of user's interest and
engagement with a virtual agent displaying shared attention behavior, by using
eye gaze and head direction information is proposed. Similarly, in
\cite{nakano2010estimating} an approach to estimate the user's conversational
engagement with a conversational agent based on analysis of gaze patterns was
implemented. Other significant work \cite{Castellano:2009} focus on feature
evaluation related to the user's non-verbal behavior, the task and the
companion's affective reactions are identified to predict the children's level
of engagement using a Bayesian model.

@@

@@

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\columnwidth]{drawSetup}
    \caption{\small \textbf{Experimental setting} Graphical representation of the set-up and the possible children focuses of attention in the scene, in red.}
    \label{drawSetup}
\end{figure}

@@

, to detect and extract the skeleton
of the subject's face and then, use the obtained features to calculate the
head pose orientation estimation (see figure \ref{system}). In this way, we can
estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
representation as it is executed as a ROS (Robot Operating System) node that can
be seen in figure \ref{rviz}.

@@

Our results for a peer-to-peer work scenarios using an approach based on RGB
camera are summarized in table \ref{tab:results} where the \textit{expected} is
the focus of attention that given the state of the system, it is likely to
attract the subject's gaze. In the same manner, the \textit{captured} is the
focus of attention our approach provides and the \textit{real}, the ground truth
of the subject's gaze direction based on manual annotations from two evaluators
with an average reliability of $ \alpha = .89 $\fixme{Is that Cronbach alpha or
Krippendorff's alpha?}.

In order to calculate the inter-rater reliability between the two measurements
(real-captured and real-expected), the overlap percentage between the two
curves\fixme{which curve?}
has been computed. A more reliable measurement in terms of time shifting between
the real and captured focuses of attention is provided using the cosine
similarity\fixme{this needs to be better justified: why not a Cronbach alpha?} also shown in table~\ref{tab:results}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{matches}
    \caption{\small \textbf{Comparison of computed focus of attention \vs ground
        truth} during a 20+ min face-to-face child-robot interaction.
        In orange (top line), the focus of attention as computed by the robot;
        in blue (bottom line), the focus of attention as manually annotated. The
    bottom graph shows agreements.}
